### 2014-03-24 reanalysis/Fadi_Towfic_generic_medicine/old_rmd/towfic_et_al_tomuch_again.rmd
figure

This document has five main parts
- Getting and formatting the data
- Reproduce some of the results to show that we are working on the same data and analysis workflow
- Remove the use of ComBat and perform an similar analysis with alternative established tools
- Estimate the error introduced by ComBat and the consequences for the conclusion of the study
- Perform a few more sanity checks to substantiate that the difference in results for the two above analysis is mainly introduced by ComBat

```{r runparameters, results='hide', echo=FALSE}
starttime = Sys.time()
debug = FALSE
downloaddata=TRUE
```

@@

```{r, results='hide', message=FALSE}
includelibs = c("Biobase", "GEOquery", "sva", "limma",
                 "preprocessCore")
lapply(includelibs, require, character.only=T)
```
The important sample information is described in Table S1 from Towfic et al. and its usage in ComBat is described briefly in the methods;
>  Each microarray’s chip designation was supplied a batch label; there were 18 batches in all. The labels for the treatments (i.e. drug product, reference standard…) were added as covariates.<cite> Towfic et al.

@@

```{r}
sampleannotation = read.table("data/sampleannotation.csv", sep="\t", header=TRUE,  stringsAsFactors=FALSE)
sampleannotation$code = make.names(sampleannotation$code)
sampleannotation$chip = as.character(sampleannotation$chip)
dimnames(sampleannotation)[[1]] = sampleannotation$code
head(sampleannotation)
```
The "covariate"-column is made based on the "code"-column, but might not match  to what were actually used for all samples. The last two columns are from the GEO deposit and reveal that 3 of the samples do not have data and will be removed from the sampleannotation.

The naming convention seem to differ slightly between the Table S1 and the text in Towfic et al. This is our interpretation of covariate labels and its corresponding name in the text.
- **DP** referred to as GA (but not GA as in table S1)
- **N** referred to as "generic"
- **M** referred to as "medium"
- **RS** referred to as "reference standard""

```{r}
# take out 3 samples that are not assign to a geoaccession. Failed QC?
sampleannotation = sampleannotation[!is.na(sampleannotation$geoaccession),] 
```
The batch/covariate design shows many batches and covariate groups.
```{r}
table(sampleannotation[, c("chip", "covariate")])
```
A look at the most interesting groups, "DP"(GA) and "N"(generic) reveals a lack of balance.
```{r}
table(sampleannotation[sampleannotation$covariate %in% c("DP", "N"), 
                       c("chip", "covariate")])
```

@@

if(downloaddata)
{
  temp <- tempfile()
  download.file(url="http://www.ncbi.nlm.nih.gov/geo/download/?acc=GSE40566&format=file&file=GSE40566%5Fnon%5Fnormalized%2Etxt%2Egz",
              destfile=temp, mode = "wb")
  rawdata <- read.table(temp, sep="\t", header=TRUE, stringsAsFactors=FALSE)
  unlink(temp)

  temp <- tempfile()
  download.file(url="http://www.ncbi.nlm.nih.gov/geo/download/?acc=GSE40566&format=file",
              destfile=temp, mode = "wb")
  tardirtemp = tempdir()
  untar(temp, exdir = tardirtemp)
  rawannotation = read.table(paste(tardirtemp ,"/GPL6887_MouseWG-6_V2_0_R3_11278593_A.txt.gz", sep=""), 
                             sep="\t", header=TRUE, stringsAsFactors=FALSE, 
                             skip=8, comment.char="", quote="", fill=TRUE)

@@

```{r}
datamatrices = list()
datamatrices[["real_raw"]] = datamatrix_raw#[1:1000,] # debug

@@

```{r reproduced_tables2, dev='png', fig.width=8, fig.height=8, fig.show='hold', tidy=FALSE}
# scatter plot correlation of the 4 columns in table_S2 vs. the reproduced numbers. String correlation.
par(mfrow=c(2, 2))
plot(table_s2$AVG_Medium, meanM[table_s2$ID], pch=".")
plot(table_s2$AVG_Reference_Standard, meanRS[table_s2$ID], pch=".")
plot(table_s2$AVG_GA, meanDP[table_s2$ID], pch=".")
plot(table_s2$AVG_generic, meanN[table_s2$ID], pch=".")
```
Original values from table S2 on X-axis, reporduced values on Y-axis. The values in Table S2 is almost reproduced.

@@

How the fold change was calculated is not explained. A common way is to take the difference of geometric means that were calculated for Table S2 above.
```{r dev='png', fig.width=8, fig.height=8, fig.show='hold', tidy=FALSE}
foldchange_DP_N = meanDP - meanN

 # make the axis equal.
lim =c( min(   c(min(table_s5$Fold_Change, na.rm=T), min(foldchange_DP_N[table_s5$Probe], na.rm=T))  ), 
        max( c(max(table_s5$Fold_Change, na.rm=T), max(foldchange_DP_N[table_s5$Probe], na.rm=T)) ))

# plot
plot(table_s5$Fold_Change, foldchange_DP_N[table_s5$Probe],
     ylim=lim, xlim=lim, pch=".", xlab="Table_S5", ylab="reproduced",
     main="FC comparison")
rm(lim)
```
The plot reveals that the reproduced fold changes are much more compressed than the ones in Table S5.
There seems to two linear relationship, one for negative FC's and another for positive FC's. They seems to have different intercepts but comparable slope.

The average expression for "GA" (DP) and "generic" (N) are also given in Table S2. The fold change calculated from those numbers compared to the ones given in Table S5 shows a similar pattern:

```{r dev='png', fig.width=8, fig.height=8, fig.show='hold', tidy=FALSE}
# table S2 has the AVG for both DP and N listed
fc_table_s2 = table_s2$AVG_GA - table_s2$AVG_generic
names(fc_table_s2) = table_s2$ID
x = table_s5$Fold_Change
y = fc_table_s2[table_s5$Probe]
lim =c( min(   c(min(x, na.rm=T), min(y, na.rm=T))  ), max( c(max(x, na.rm=T), max(y, na.rm=T)) ))
plot(x, y , ylim=lim, xlim=lim, pch="." ,  xlab="Table_S5", ylab="Table_S2",
     main="FC comparison, given values.") #  almost perfect match with offset.
rm(x, y, lim)
```

@@

thiscolors = c("black", "red")
hist(foldchange_DP_N, breaks=100, freq=F, border=thiscolors[1], main="Reproduced FC for DP vs. N")
hist(foldchange_DP_N[table_s5$Probe], breaks=100, add=T, freq=F, border=thiscolors[2])
legend("topright", legend=c("All probes", "Table S5 probes"), text.col=thiscolors)
```

The reproduced fold changes for the probes listed in Table S5 are clearly more extreme than the rest.

The next test is a test for differentially expressed genes using ANOVA.
>  First, we computed the statistical significance of differential expression between treatments using the Analysis of Variance (ANOVA) method for each probe [21]...<cite> Towfic et al.

>>>>

This document has five main parts
- Getting and formatting the data
- Reproduce some of the results to show that we are working on the same data and analysis workflow
- Remove the use of ComBat and perform a similar analysis with alternative established tools
- Estimate the error introduced by ComBat and the consequences for the conclusion of the study
- Perform a few more sanity checks to substantiate that the difference in results for the two above analysis is mainly introduced by ComBat

```{r runparameters, results='hide', echo=FALSE}
starttime = Sys.time()
debug = TRUE
downloaddata=TRUE
```

```{r, results='hide', message=FALSE}
includelibs = c("Biobase", "GEOquery", "sva", "limma",
                 "preprocessCore")
lapply(includelibs, require, character.only=T)
```

@@

The important sample information is described in Table S1 from Towfic et al. and its usage in ComBat is described briefly in the methods;
>  Each microarray’s chip designation was supplied a batch label; there were 18 batches in all. The labels for the treatments (i.e. drug product, reference standard…) were added as covariates.<cite> Towfic et al.

@@

```{r}
sampleannotation = read.table("data/sampleannotation.csv", sep="\t", header=TRUE,  stringsAsFactors=FALSE)
sampleannotation$code = make.names(sampleannotation$code)
sampleannotation$chip = as.character(sampleannotation$chip)
dimnames(sampleannotation)[[1]] = sampleannotation$code
head(sampleannotation)
```
The "covariate"-column is made based on the "code"-column, but might not match 100% to what were actually used for all samples. The last two columns are from the GEO deposit and reveal the samples that do not have data, which we will remove from the data shortly.

The naming convention seem to differ slightly between the Table S1 and the text in Towfic et al. This is our interpretation of the main covariate labels and its corresponding name in the text.
- **DP** referred to as GA (but not GA as in table S1)
- **N** referred to as "generic"
- **M** referred to as "medium"
- **RS** referred to as "reference standard""

```{r}
# take out 3 samples that are not assign to a geoaccession. Failed QC?
sampleannotation = sampleannotation[!is.na(sampleannotation$geoaccession),] 
```
The batch/covariate design shows many batches and covariate groups.
```{r}
table(sampleannotation[, c("chip", "covariate")])
```
A look at the primary comparison, "DP"(GA) and "N"(generic) reveals a lack of balance.
```{r}
table(sampleannotation[sampleannotation$covariate %in% c("DP", "N"), 
                       c("chip", "covariate")])
```

@@

if(downloaddata)
{
  temp = tempfile()
  download.file(url="http://www.ncbi.nlm.nih.gov/geo/download/?acc=GSE40566&format=file&file=GSE40566%5Fnon%5Fnormalized%2Etxt%2Egz",
              destfile=temp, mode = "wb")
  rawdata = read.table(temp, sep="\t", header=TRUE, stringsAsFactors=FALSE)
  unlink(temp)

  temp = tempfile()
  download.file(url="http://www.ncbi.nlm.nih.gov/geo/download/?acc=GSE40566&format=file",
              destfile=temp, mode = "wb")
  tardirtemp = tempdir()
  untar(temp, exdir = tardirtemp)
  rawannotation = read.table(paste(tardirtemp ,"/GPL6887_MouseWG-6_V2_0_R3_11278593_A.txt.gz", sep=""), 
                             sep="\t", header=TRUE, stringsAsFactors=FALSE, 
                             skip=8, comment.char="", quote="", fill=TRUE)

@@

```{r}
datamatrices = list()
datamatrices[["real_raw"]] = datamatrix_raw#[1:1000,] # debug
if(debug)
  datamatrices[["real_raw"]] = datamatrix_raw[1:1000,]

@@

```{r reproduced_tables2, dev='png', fig.width=8, fig.height=8, fig.show='hold', tidy=FALSE}
# scatter plot correlation of the 4 columns in table_S2 vs. the reproduced numbers. String correlation.
par(mfrow=c(2, 2))
plot(table_s2$AVG_Medium, meanM[table_s2$ID], pch=".", na.rm=T)
plot(table_s2$AVG_Reference_Standard, meanRS[table_s2$ID], pch=".", na.rm=T)
plot(table_s2$AVG_GA, meanDP[table_s2$ID], pch=".", na.rm=T)
plot(table_s2$AVG_generic, meanN[table_s2$ID], pch=".", na.rm=T)
```
Original values from table S2 on X-axis, reporduced values on Y-axis. The values in Table S2 is almost reproduced.

@@

How the fold change was calculated is not explained. A common way is to take the difference of geometric means that were calculated for Table S2 above.
```{r}
foldchange_DP_N = meanDP - meanN
```

These reproduced foldchanges can be compared for the probes listed in table S5.

@@

thiscolors = c("black", "red")
hist(foldchange_DP_N, breaks=100, freq=F, border=thiscolors[1], main="Reproduced FC for DP vs. N")
hist(foldchange_DP_N[table_s5$Probe], breaks=100, add=T, freq=F, border=thiscolors[2])
legend("topright", legend=c("All probes", "Table S5 probes"), text.col=thiscolors)
```
The first plot reveals that the reproduced fold changes are much more compressed than the ones in Table S5.
There seems to two linear relationship, one for negative FC's and another for positive FC's. They seems to have different intercepts but comparable slope. The second plot shows that the reproduced fold changes for the probes listed in Table S5 are clearly more extreme than the rest.

The fold changes in table 5S do not match the average expression values listed in table S2 either as a the same plot will show;

```{r dev='png', fig.width=8, fig.height=8, fig.show='hold', tidy=FALSE}
# table S2 has the AVG for both DP and N listed
fc_table_s2 = table_s2$AVG_GA - table_s2$AVG_generic
names(fc_table_s2) = table_s2$ID
x = table_s5$Fold_Change
y = fc_table_s2[table_s5$Probe]
lim =c( min(   c(min(x, na.rm=T), min(y, na.rm=T))  ), max( c(max(x, na.rm=T), max(y, na.rm=T)) ))
plot(x, y , ylim=lim, xlim=lim, pch="." ,  xlab="Table_S5", ylab="Table_S2",
     main="FC comparison, given values.") #  almost perfect match with offset.
rm(x, y, lim)
```
There seems to be some inconsistancy between between the fold changes in table S5 and the reproduced ones, however it is the conficende i.e p-values that are affected by the use of ComBat. The rest of table S5 is p-values and we will ty and reproduce some of them.

ANOVA is The first test for differentially expressed gene.
>  First, we computed the statistical significance of differential expression between treatments using the Analysis of Variance (ANOVA) method for each probe [21]...<cite> Towfic et al.
